## `r img_label`

`r n` rows by `r p` columns, values running from
`r sprintf("%.3g to %.3g", dsc$range_x[1], dsc$range_x[2])`.
The first component carries
`r sprintf("%.2f%%", 100 * dsc$var_share_1)` of `trace(X'X)`, and the first ten
carry `r sprintf("%.2f%%", 100 * dsc$var_share_10)`.

### The original, then k = 1 through 10

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: !expr grid_h
op <- par(mfrow = c(grid_nr, grid_nc), oma = c(0, 0, 0, 0))
rk <- sum(dcm$lambda > max(dcm$lambda) * 1e-12)
show_image(X, zlim = zl, cex_main = cexm,
           main = if (grid_nc >= 8) sprintf("orig (rank %d)", rk)
                  else sprintf("original (rank %d)", rk))
for (k in 1:10) {
  show_image(pca_compress(X, k, decomp = dcm)$approximation,
             main = sprintf("k = %d", k), zlim = zl, cex_main = cexm)
}
for (i in seq_len(grid_nr * grid_nc - 11)) plot.new()
par(op)
```

Every panel is drawn on the original image's grey scale, so what you see is a
like-for-like comparison. A side effect is that where a reconstruction
overshoots past the original's brightest pixel, it clips to pure white.

### Going past k = 10

Ten components isn't always enough to judge where to stop, so here is the same
image at a wider spread of $k$, finishing at $k = p$, which is exact.

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: !expr grid2_h
op <- par(mfrow = c(grid2_nr, grid2_nc), oma = c(0, 0, 0, 0))
for (k in k_more) {
  show_image(pca_compress(X, k, decomp = dcm)$approximation,
             main = sprintf("k = %d  (err %.3f)", k, err[k] / normX),
             zlim = zl, cex_main = cexm)
}
for (i in seq_len(grid2_nr * grid2_nc - length(k_more))) plot.new()
par(op)
```

Titles give $\text{Error}(k)$ as a fraction of $\lVert X \rVert_F$.

### Error at every k from 1 to `r p`

```{r}
#| echo: false
#| fig-width: 8.5
#| fig-height: 4.6
plot_error_curve(err, mark = k_choice,
                 main = sprintf("%s — Frobenius error vs k", img_label))
```

```{r}
#| echo: false
knitr::kable(
  data.frame(
    k                = k_tab,
    `Error(k)`       = round(err[k_tab], 2),
    `% of Error(1)`  = sprintf("%.1f%%", 100 * err[k_tab] / err[1]),
    `rel. residual`  = sprintf("%.4f", err[k_tab] / normX),
    `storage vs raw` = sprintf("%.1f%%",
                        100 * sapply(k_tab, function(k)
                          pca_compression_ratio(n, p, k)$ratio)),
    check.names = FALSE
  ),
  caption = sprintf("%s — error and storage cost at selected k. The residual column is Error(k) divided by ||X||_F.", img_label)
)
```

### The first eigenvector and the first score vector

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4.2
plot_u1_z1(dsc$u1, dsc$z1)
```

Numbers behind those two curves, which the discussion below refers back to:

```{r}
#| echo: false
cat(sprintf(
"u1 : %.1f%% of its %d entries are positive; range [%.4f, %.4f];
      largest-magnitude entry at column %d; dispersion sd/mean|u1| = %.3f
z1 : range [%.1f, %.1f]; largest magnitude at row %d; minimum at row %d
      (rows are numbered from the top of the image)\n",
100 * dsc$u1_sign_frac, p, dsc$u1_range[1], dsc$u1_range[2],
dsc$u1_peak_col, dsc$u1_flatness,
dsc$z1_range[1], dsc$z1_range[2], dsc$z1_peak_row, dsc$z1_min_row))
```

`r interp_md`
