---
title: "Image Compression with Principal Components"
subtitle: "Rank-k approximation on the uncentered cross product X'X"
author: "Kyle Fraund"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    code-tools: false
    fig-align: center
    tbl-cap-location: bottom
knitr:
  opts_chunk:
    comment: ""
    dev: png
    dpi: 130
---

```{r}
#| label: setup
#| include: false
options(knitr.duplicate.label = "allow")
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

source(file.path("R", "pca_compress.R"))
source(file.path("R", "plot_helpers.R"))
library(readr)

k_choice_by_image <- tryCatch(
  source(file.path("config", "k_choices.R"), local = TRUE)$value,
  error = function(e) list()
)

MAX_P <- 1200L
```

```{r}
#| label: load-data
#| include: false
csv_files <- sort(Sys.glob(file.path("data", "*.csv")))

if (length(csv_files) == 0L) {
  stop("No CSV files found in data/. Put the image CSVs there ",
       "(image1.csv, image2.csv, ...) and re-render.", call. = FALSE)
}

read_image_csv <- function(path) {
  df <- readr::read_csv(path, show_col_types = FALSE, progress = FALSE)
  m  <- data.matrix(df)

  if (!is.numeric(m))        stop(basename(path), ": not all columns are numeric.", call. = FALSE)
  if (anyNA(m))              stop(basename(path), ": contains ", sum(is.na(m)),
                                  " missing values.", call. = FALSE)
  if (nrow(m) < 2 || ncol(m) < 2) stop(basename(path), ": too small to be an image.", call. = FALSE)
  if (ncol(m) > MAX_P)       stop(basename(path), ": p = ", ncol(m), " exceeds MAX_P = ",
                                  MAX_P, ".", call. = FALSE)
  dimnames(m) <- NULL
  m
}

images <- lapply(csv_files, read_image_csv)
names(images) <- tools::file_path_sans_ext(basename(csv_files))
```

## What this page is

Each of the four images below is a numeric matrix $X$, one entry per pixel,
$n$ rows by $p$ columns. I decompose each one without centering or scaling,

$$V = X'X, \qquad V = U \Lambda U', \qquad Z = XU, \qquad X = ZU'$$

and then throw away all but the first $k$ components:

$$X_{(k)} = z_1u_1' + z_2u_2' + \cdots + z_ku_k',
\qquad \text{Error}(k) = \lVert X - X_{(k)} \rVert_F .$$

Because $U$ is orthonormal, $UU' = I$ and $X = ZU'$ holds exactly, so setting
$k = p$ hands back the original image untouched. Any smaller $k$ drops the
directions that contribute least to $\operatorname{trace}(X'X)$, and the whole
question is how many of them you can drop before you notice.

It matters that nothing is centered. `prcomp()` subtracts column means by
default, and once it does, $ZU'$ reconstructs $X$ minus those means rather than
$X$ itself, so the reconstruction is no longer the image. Everything here goes
through `eigen(crossprod(x), symmetric = TRUE)` instead.

A note on the data, since it changes how the plots read: these pixel values
aren't on a 0-255 scale. They're roughly zero-centered floats, some negative.
That means $X'X$ is picking up energy, or distance from mid-grey, rather than
brightness, which is why the first eigenvector doesn't always look like an
average column the way you might expect.

## Part 1: the compression function

Here is `R/pca_compress.R` in full.

```{r}
#| label: echo-source
#| echo: false
#| results: asis
cat("```r\n", paste(readLines(file.path("R", "pca_compress.R")), collapse = "\n"),
    "\n```\n", sep = "")
```

`pca_compress(x, k)` is the piece the assignment asks for. It checks its inputs,
decomposes $X'X$, truncates to $k$ components and returns both the rank-$k$
matrix and its Frobenius error. Input validation rejects a non-numeric or
non-matrix `x`, a `k` that isn't a single whole number, and any `k` outside
$1 \le k \le p$. The approximation itself is the one-line matrix product
`Z[, 1:k] %*% t(U[, 1:k])`, with `drop = FALSE` so that $k = 1$ stays a matrix
instead of collapsing to a vector.

The other three functions exist for speed and convenience:

- `pca_decomp(x)` does the eigendecomposition and returns `U`, `lambda` and
  `Z = XU`. You can hand the result back to `pca_compress()` through its
  `decomp` argument, which matters here because this page needs an error for
  every $k$ from 1 to $p$ and redoing an $O(p^3)$ eigendecomposition a few
  hundred times would be silly. `crossprod(x)` is used rather than
  `t(x) %*% x` because it's about twice as fast, and `symmetric = TRUE` forces
  the real-symmetric LAPACK routine, which guarantees a genuinely orthonormal
  $U$ and returns the eigenvalues already sorted largest first.
- `pca_error_curve(x)` walks $k$ from 1 to $p$ accumulating rank-one updates,
  $X_{(k)} = X_{(k-1)} + z_ku_k'$. Rebuilding each approximation from scratch
  would cost $O(np^3)$ across the sweep; this way it's $O(np^2)$. These are real
  `norm(..., type = "F")` values on the residual, not the
  $\sqrt{\sum \lambda}$ shortcut, which is the point of the second check below.
- `pca_compression_ratio(n, p, k)` is the storage arithmetic: $(n+p)k$ numbers
  kept against $np$ in the raw image.

### Does it actually work?

Three things have to be true, and all three are checked on the real data rather
than asserted:

```{r}
#| label: verification
#| echo: false
verify_one <- function(nm, X) {
  d   <- pca_decomp(X)
  p   <- ncol(X)
  nX  <- norm(X, type = "F")
  err <- pca_error_curve(X, decomp = d)

  tail_sum <- rev(cumsum(rev(d$lambda)))
  theory   <- c(sqrt(tail_sum[-1]), 0)

  cat(sprintf("=== %s  (n = %d, p = %d) ===\n", nm, nrow(X), p))
  cat(sprintf("1. Error(k = p) = %.4e   vs  ||X||_F = %.2f   -> relative %.2e\n",
              err[p], nX, err[p] / nX))
  cat(sprintf("2. max_k |Error(k) - sqrt(sum lambda_{k+1..p})| = %.4e\n",
              max(abs(err - theory))))
  cat(sprintf("   e.g. k = 5 : Error = %.6f , sqrt(tail) = %.6f\n",
              err[min(5, p)], theory[min(5, p)]))
  cat(sprintf("3. Error monotone non-increasing: %s  (largest step up = %.3e)\n\n",
              all(diff(err) <= 1e-8), max(diff(err))))
  invisible(NULL)
}
invisible(verify_one(names(images)[1], images[[1]]))
```

Keeping every component reproduces the image to fifteen significant figures, the
Frobenius error agrees with the square root of the discarded eigenvalues to
within floating-point noise, and the error never goes up as $k$ grows. There's
also a separate test file, `tests/test_pca_compress.R`, that runs seventeen
checks on throwaway matrices, including the input validation and both the
$n < p$ and $n > p$ cases.

## Part 2: the images

```{r}
#| label: per-image
#| echo: false
#| results: asis
tmpl <- readLines("_image_template.qmd")

out <- character(0)
for (i in seq_along(images)) {
  X         <- images[[i]]
  img_name  <- names(images)[i]
  img_label <- img_name
  n <- nrow(X); p <- ncol(X)

  dcm   <- pca_decomp(X)
  err   <- pca_error_curve(X, decomp = dcm)
  normX <- norm(X, type = "F")
  dsc   <- describe_pc1(X, dcm, err)
  zl    <- range(X, finite = TRUE)

  k_choice <- k_choice_by_image[[img_name]]
  k_tab    <- unique(pmin(c(1, 2, 3, 5, 10, 20, 40, 80, p), p))

  aspect <- n / p
  if (aspect >= 3) {
    grid_nr <- 1;  grid_nc <- 11
  } else if (aspect <= 1/3) {
    grid_nr <- 11; grid_nc <- 1
  } else {
    grid_nr <- 3;  grid_nc <- 4
  }
  grid_h <- max(4, min(22, grid_nr * (10 / grid_nc) * aspect + 0.45 * grid_nr))
  cexm   <- if (grid_nc >= 8) 0.8 else 1

  k_more    <- unique(pmin(c(15, 20, 30, 40, 60, p), p))
  grid2_nc  <- if (grid_nr == 1) length(k_more) else min(3, length(k_more))
  grid2_nr  <- ceiling(length(k_more) / grid2_nc)
  grid2_h   <- max(3.5, min(22, grid2_nr * (10 / grid2_nc) * aspect + 0.45 * grid2_nr))

  interp_path <- file.path("interpretations", paste0(img_name, ".md"))
  interp_md <- if (file.exists(interp_path)) {
    paste(readLines(interp_path, warn = FALSE), collapse = "\n")
  } else {
    paste0("::: {.callout-important title=\"Interpretation not written yet\"}\n",
           "No `", interp_path, "` found.\n:::")
  }

  out <- c(out, knitr::knit_child(text = tmpl, envir = environment(),
                                  quiet = TRUE))
}
cat(paste(out, collapse = "\n\n"))
```

## Rebuilding this page

```bash
Rscript tests/test_pca_compress.R
quarto render
```
